\pkg{Rgbp}: Bayesian Hierarchical Modeling and Frequentist Method Check

### 2013-04-12 jss/article.tex
picture

\subsection{Binomial: 18 Baseball Data}
This dataset contains information about batting averages of 18 major league players through their first 45 official at-bats of the 1970 season (Efron and Morris, 1975). Also, it has two covariates, League and Position, showing in which league and in which position each player was. In this paper, we will use Position for a tutorial purpose. For convenience, we transform this variable into a binary indicator, which is 1 if a player was a outfielder and 0 otherwise. The code below shows a way to make inputs. If we have more than one covariate, for example, \code{x1} and \code{x2}, then `\code{R> x <- cbind(x1, x2)}' will be the right input of the \code{gbp} function.
\begin{CodeChunk}
\begin{CodeInput}
R> z <- c(18, 17, 16, 15, 14, 14, 13, 12, 11, 11, 10, 10, 10, 10, 10,  9,  8,  7)
R> n <- c(45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45)
R> x <- c( 1,  1,  1,  1,  1,  0,  0,  0,  0,  1,  0,  0,  0,  1,  1,  0,  0,  0) 
\end{CodeInput}
\end{CodeChunk}
The data indicate that independent Binomial distribution is appropriate for each player's number of hits among 45 at-bats conditioning on the unknown parameter $p_{j}$: $z_{j}\vert p_{j}\stackrel{ind}{\sim} \textrm{Binomial}(n_{j}, p_{j}), ~j=1, \ldots, 18$. This dataset is also a part of the package and can be called on \proglang{R} by `\code{R> data(baseball)}'.

\section[Multi-level Structure]{Multi-level Structure}
Our multi-level models, also called conditionally independent hierarchical models (Kass and Steffey, 1989), is a very powerful tool for exploring the hierarchical sturucture in data. For example, we can think about a district-level hierarchy (bigger population) for 8 schools, the state-level hierarchy for 31 hospitals, and the position-level hierarchy for 18 baseball players. \code{gbp}, one of functions in \pkg{Rgbp}, fits such a hierarchical model whose first-level hierarchy has a distribution of observed data and second-level (bigger population hierarchy) has a conjugate prior distribution on the first-level parameter. Users can determine one of three types of multi-level models, such as Normal-Normal, Poisson-Gamma, and Binomial-Beta, based on their datasets. 
\\

@@

where $\mu_{0j} =x^{T}_{j}\beta,~j=1, \ldots, 8$, $x_{j}$ is $j$-th school's covariate vector ($1\times m$), and $m$ is the number of regression coefficients which is just 1 for an intercept term if there is no covariate. For reference, a paramter with a zero subscript, $\mu_{0j}$, represents a mean parameter of the prior (second-level) distribution. Also,  Based on this conjugate prior distribution, it is easy to derive corresponding posterior distribution. 
\begin{equation}
\mu_{j}\vert \textbf{y}, \beta, A \stackrel{ind}{\sim}\textrm{Normal}(~(1-B_{j})y_{j} + B_{j}\mu_{0j},~(1-B_{j})\sigma^{2}_{j}),
\end{equation}
where $B_{j}\equiv\frac{\sigma^{2}_{j}}{\sigma^{2}_{j} + A},~j=1, \ldots, 8$, is called a shrinkages.

\subsection[Poisson-Gamma]{Poisson-Gamma}
\code{gbp} is also able to build a Poisson-Gamma multi-level model on the 31 hospital data. Note that a constant multiplied to the notation representing Gamma distribution below is a scale and a square bracket below indicates [mean, variance] of distribution.
\begin{equation}
z_{j}\vert \lambda_{j} \stackrel{ind}{\sim}\textrm{Poisson}(n_{j}\lambda_{j}),
\end{equation}
\begin{equation}
\lambda_{j}\vert \beta, r\stackrel{ind}{\sim}\frac{1}{r}\textrm{Gamma}(\lambda_{0j}r)\sim \textrm{Gamma}[\lambda_{0j}, ~\frac{\lambda_{0j}}{r}],
\end{equation}

where $\log(\lambda_{0j}) =x_{j}'\beta$, and $j=1, \ldots, 31$. Immediate posterior distribution of this Poisson-Gamma model is
\begin{equation}
\lambda_{j}\vert \textbf{z}, \beta, r \stackrel{ind}{\sim}\frac{1}{r + n_{j}}\textrm{Gamma}(r\lambda_{0j} + z_{j})\sim\textrm{Gamma}[\lambda^{\ast}_{j},~\frac{\lambda^{\ast}_{j}}{r+n_{j}}],
\end{equation}

where $\lambda^{\ast}_{j} \equiv (1-B_{j})(\frac{z_{j}}{n_{j}}) + B_{j}\lambda_{0j},~\textrm{and}~ B_{j}\equiv \frac{r}{r+n_{j}},~ j=1, \ldots, 31$.

\subsection[Binomial-Beta]{Binomial-Beta}
Binomial-Beta hierarchical model is the last model that \code{gbp} can fit. Again, a square bracket below indicates [mean, variance] of distribution.
\begin{equation}
z_{j} \vert p_{0j}\stackrel{ind}{\sim}\textrm{Binomial}(n_{j}, ~p_{j}),
\end{equation}
\begin{equation}
p_{j} \vert \beta, r\stackrel{ind}{\sim}\textrm{Beta}(rp_{0j},~ r(1-p_{0j}))\sim \textrm{Beta}[p_{0j}, ~\frac{p_{0j}(1-p_{0j})}{r + 1}],
\end{equation}

where $\log(\frac{p_{0j}}{1-p_{0j}}) =x_{j}'\beta~\textrm{and}~j=1, \ldots, 18$. Then posterior distribution is
\begin{equation}
p_{j}\vert \textbf{z}, \beta, r \stackrel{ind}{\sim}\textrm{Beta}(rp_{0j}+z_{j},~r(1-p_{0j})+n_{j}-z_{j})\sim\textrm{Beta}\bigg[p^{\ast}_{j},~ \frac{p^{\ast}_{j}(1-p^{\ast}_{j})}{r+n_{j}+1}\bigg],
\end{equation}
where $p^{\ast}_{j}\equiv(1-B_{j})(\frac{z_{j}}{n_{j}})+B_{j}p_{0j},~ \textrm{and}~ B_{j}\equiv\frac{r}{r+n_{j}}$, $j=1,\ldots,18$.

\subsection[Hyper-prior Distribution]{Hyper-prior Distribution}
Hyper-prior distribution indicates a distribution of the second-level parameters, which plays an important role in deriving a full posterior distribution of all the parameters. \code{gbp} sets a non-informative distribution on second-level parameters to let the data speak more about their estimation.
\begin{equation}
\beta \sim \textrm{flat on}~ \mathbf{R}^{m},~~A ~(\textrm{or}~ \frac{1}{r})\sim \textrm{Unif}(0, \infty),
\end{equation}

where $m$ is the number of regression coefficients. For $\beta$, it is reasonable choice to take flat (non-informative) distribution because the information about location gets informative as number of groups increases. Another flat prior on the hyper-prameter $A$ (or $1/r$) that we suggest here will guarantee a posterior propriety under the moderate conditions and will give users a good repeated sampling property.
\\

@@

In this case, one way is to obtain an MLE of the variance component with its asymptotic Normal distribution and to use Delta method for asymptotic Normal distribution of shrinkage. But is the Normal approximation a good approximation for shrinkage that takes on values between 0 and 1?
\\

@@

\section[Reference]{Reference}
Christiansen, C. and Morris, C. (1996). ``Fitting and Checking a Two-Level Poisson Model: Modeling Patient Mortality Rates in Heart Transplant Patients,'' in \emph{Bayesian Biostatistics}, eds. D. Berry and D. Stangl, New York: Marcel Dekker, pp. 467-561.
\\

Efron, B. and Morris, C. (1975). ``Data Analysis Using Stein's Estimator and its Generalizations.'' \emph{Journal of the American Statistical Association}. \textbf{70}. 311-319.
\\

Kass, R. and Steffey, D. (1989). ``Approximate Bayesian Inference in Conditionally Independent Hierarchical Models (Parametric
Empirical Bayes Models).'' \emph{Journal of the American Statistical Association}. \textbf{84}. 717-726.
\\

Morris, C. and Tang, R. (2011). ``Estimating Random Effects via Adjustment for Density Maximization.'' \emph{Statistical Science}. \textbf{26}. 271-287.
\\

Morris, C. and Lysy, M. (2012). ``Shrinkage Estimation in Multilevel Normal Models.'' \emph{Statistical Science}. \textbf{27}. 115-134.
\\

Rubin, D. B. (1981). ``Estimation in parallel randomized
  experiments.'' \emph{Journal of Educational Statistics}. \textbf{6}. 377-401.
\\

>>>>

\subsection{Binomial: 18 Baseball Data}
This dataset contains information about batting averages of 18 major league players through their first 45 official at-bats of the 1970 season (Efron and Morris, 1975). Also, it has two covariates, League and Position, showing in which league and in which position each player was playing. In this paper, we will use Position for a tutorial purpose. For convenience, we transform this variable into a binary indicator, which is 1 if a player was a outfielder and 0 otherwise. The code below shows a way to make inputs. If we have more than one covariate, for example, \code{x1} and \code{x2}, then `\code{R> x <- cbind(x1, x2)}' will be the right input of the \code{gbp} function.
\begin{CodeChunk}
\begin{CodeInput}
R> z <- c(18, 17, 16, 15, 14, 14, 13, 12, 11, 11, 10, 10, 10, 10, 10,  9,  8,  7)
R> n <- c(45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45)
R> x <- c( 1,  1,  1,  1,  1,  0,  0,  0,  0,  1,  0,  0,  0,  1,  1,  0,  0,  0) 
\end{CodeInput}
\end{CodeChunk}
The data indicate that independent Binomial distribution is appropriate for each player's number of hits among 45 at-bats conditioning on the unknown parameter $p_{j}$: $z_{j}\vert p_{j}\stackrel{ind}{\sim} \textrm{Binomial}(n_{j}, p_{j}), ~j=1, \ldots, 18$. This dataset is also a part of the package and can be called on \proglang{R} by `\code{R> data(baseball)}'.

\section[Multi-level Structure]{Multi-level Structure}
Our multi-level model, also called a conditionally independent hierarchical model (Kass and Steffey, 1989), is a very powerful tool for exploring the hierarchical sturucture in data. For example, we can think about a district-level hierarchy (bigger population) for 8 schools, the state-level hierarchy for 31 hospitals, and the position-level hierarchy for 18 baseball players. \code{gbp}, one of functions in \pkg{Rgbp}, fits such a hierarchical model whose first-level hierarchy has a distribution of observed data and second-level (bigger population hierarchy) has a conjugate prior distribution on the first-level parameter. Users can determine one of three types of multi-level models, such as Normal-Normal, Poisson-Gamma, and Binomial-Beta, based on their datasets. 
\\

@@

where $\mu_{0j} =x^{T}_{j}\beta,~j=1, \ldots, 8$, $x_{j}$ is $j$-th school's covariate vector ($m\times 1$), and $m$ is the number of regression coefficients. Note that if there is no covariate then $m$ is 1 for an intercept term, making $\mu_{0j}=\mu_{0}=\beta_{0}$ for all $j$. For reference, a paramter with a zero subscript, $\mu_{0j}$, represents a mean parameter of the prior (second-level) distribution. Also,  Based on this conjugate prior distribution, it is easy to derive corresponding posterior distribution.
\begin{equation}
\mu_{j}\vert \textbf{y}, \beta, A \stackrel{ind}{\sim}\textrm{Normal}(~(1-B_{j})y_{j} + B_{j}\mu_{0j},~(1-B_{j})\sigma^{2}_{j}),
\end{equation}
where $B_{j}\equiv\frac{\sigma^{2}_{j}}{\sigma^{2}_{j} + A},~j=1, \ldots, 8$, is called a shrinkages.

\subsection[Poisson-Gamma]{Poisson-Gamma}
\code{gbp} is also able to build a Poisson-Gamma multi-level model on the 31 hospital data. Note that a constant multiplied to the notation representing Gamma distribution below is a scale and a square bracket below indicates [mean, variance] of distribution. And for notational consistency, let's define $y_{j}\equiv \frac{z_{j}}{n_{j}}$ for all $j$.
\begin{equation}
z_{j}\vert \lambda_{j} \stackrel{ind}{\sim}\textrm{Poisson}(n_{j}\lambda_{j}),
\end{equation}
\begin{equation}
\lambda_{j}\vert \beta, r\stackrel{ind}{\sim}\frac{1}{r}\textrm{Gamma}(\lambda_{0j}r)\sim \textrm{Gamma}[\lambda_{0j}, ~\frac{\lambda_{0j}}{r}],
\end{equation}

where $\log(\lambda_{0j}) =x_{j}'\beta$, and $j=1, \ldots, 31$. Immediate posterior distribution of this Poisson-Gamma model is
\begin{equation}
\lambda_{j}\vert \textbf{z}, \beta, r \stackrel{ind}{\sim}\frac{1}{r + n_{j}}\textrm{Gamma}(r\lambda_{0j} + n_{j}y_{j})\sim\textrm{Gamma}[\lambda^{\ast}_{j},~\frac{\lambda^{\ast}_{j}}{r+n_{j}}],
\end{equation}

where $\lambda^{\ast}_{j} \equiv (1-B_{j})y_{j} + B_{j}\lambda_{0j}$,  $B_{j}\equiv \frac{r}{r+n_{j}}$, and $y_{j}\equiv \frac{z_{j}}{n_{j}}$, $j=1, \ldots, 31$.

\subsection[Binomial-Beta]{Binomial-Beta}
Binomial-Beta hierarchical model is the last model that \code{gbp} can fit. Again, a square bracket below indicates [mean, variance] of distribution.
\begin{equation}
z_{j} \vert p_{j}\stackrel{ind}{\sim}\textrm{Binomial}(n_{j}, ~p_{j}),
\end{equation}
\begin{equation}
p_{j} \vert \beta, r\stackrel{ind}{\sim}\textrm{Beta}(rp_{0j},~ r(1-p_{0j}))\sim \textrm{Beta}[p_{0j}, ~\frac{p_{0j}(1-p_{0j})}{r + 1}],
\end{equation}

where $\log(\frac{p_{0j}}{1-p_{0j}}) =x_{j}'\beta~\textrm{and}~j=1, \ldots, 18$. Then posterior distribution is
\begin{equation}
p_{j}\vert \textbf{z}, \beta, r \stackrel{ind}{\sim}\textrm{Beta}(rp_{0j}+n_{j}y_{j},~r(1-p_{0j})+n_{j}(1-y_{j})\sim\textrm{Beta}\bigg[p^{\ast}_{j},~ \frac{p^{\ast}_{j}(1-p^{\ast}_{j})}{r+n_{j}+1}\bigg],
\end{equation}
where $p^{\ast}_{j}\equiv(1-B_{j})y_{j}+B_{j}p_{0j}$, $B_{j}\equiv\frac{r}{r+n_{j}}$, and $y_{j}\equiv\frac{z_{j}}{n_{j}}$, $j=1,\ldots,18$.

\subsection[Hyper-prior Distribution]{Hyper-prior Distribution}
Hyper-prior distribution indicates a distribution of the second-level parameters, which plays an important role in deriving a full posterior distribution of all the parameters. \code{gbp} sets a non-informative distribution on second-level parameters to let the data speak more about their estimation.
\begin{equation}
\beta \sim \textrm{Uniform on}~ \mathbf{R}^{m},~~A ~(\textrm{or}~ \frac{1}{r})\sim \textrm{Uniform}(0, \infty),
\end{equation}

where $m$ is the number of regression coefficients. For $\beta$, it is reasonable choice to take flat (non-informative) distribution because information about the location gets informative as number of groups increases. Another flat prior on the hyper-prameter $A$ (or $1/r$) that we suggest here will guarantee a posterior propriety under the moderate conditions and will give users a good repeated sampling property.
\\

@@

In this case, one way is to obtain an MLE of the variance component with its asymptotic Normal distribution and then to use Delta method for asymptotic Normal distribution of shrinkage. But is the Normal approximation a good approximation for shrinkage that takes on a value between 0 and 1?
\\

@@

\section[Reference]{Reference}
1. Christiansen, C. and Morris, C. (1996). ``Fitting and Checking a Two-Level Poisson Model: Modeling Patient Mortality Rates in Heart Transplant Patients,'' in \emph{Bayesian Biostatistics}, eds. D. Berry and D. Stangl, New York: Marcel Dekker, pp. 467-561.
\\

2. Efron, B. and Morris, C. (1975). ``Data Analysis Using Stein's Estimator and its Generalizations.'' \emph{Journal of the American Statistical Association}. \textbf{70}. 311-319.
\\

3. Kass, R. and Steffey, D. (1989). ``Approximate Bayesian Inference in Conditionally Independent Hierarchical Models (Parametric
Empirical Bayes Models).'' \emph{Journal of the American Statistical Association}. \textbf{84}. 717-726.
\\

4. Morris, C. and Tang, R. (2011). ``Estimating Random Effects via Adjustment for Density Maximization.'' \emph{Statistical Science}. \textbf{26}. 271-287.
\\

5. Morris, C. and Lysy, M. (2012). ``Shrinkage Estimation in Multilevel Normal Models.'' \emph{Statistical Science}. \textbf{27}. 115-134.
\\

6. Rubin, D. B. (1981). ``Estimation in parallel randomized
  experiments.'' \emph{Journal of Educational Statistics}. \textbf{6}. 377-401.
\\
